Cartesian Closed Categories

### 2017-09-21 papers/talk-8/CCC.tex

\end{definition}
Since products and exponentials are given in the definition, the remainung task is, whether
 $$\textbf{HA}(a \land b, c) \cong \textbf{HA}(a, b \Rightarrow c)$$ holds.
\\
Heyting Algbras are partially ordered sets, so as in the example for \textbf{posets}, the monotonicity constraint must hold for exponentials $b^a= a \Rightarrow q$.
In the following, exponentation is written as $a \Rightarrow b = \neg a \lor b$, when necessary.
\\
So the monotonicity constraint imposes $$ a \Rightarrow b \leq b \Rightarrow c \ iff \ a \leq b \ and \ b \leq c.$$
Hence $$a \Rightarrow b  \leq a \Rightarrow b$$
$$ a \Rightarrow b \land \a \leq b$$ by the condition $a \land b \leq c$ iff $a \leq b \Rightarrow c$, which is the evaluation arrow, and it is monotone
$$ a \Rightarrow b \land a = (\neg a \lor b) \land a$$
$$ = (\neg a \land a) \lor (b \land a)$$
$$ = 0 \lor b \land a \leq b$$ what always holds.

@@

   \\ For the second axiom, choose $f:A\to B^B$ as and $p:=1_A$.Let $\mu := \lambda^{-1}$
   $$f \circ 1_A : A \to A \to B^B$$
   the transpose has type
   $$\lambda^{-1}: A\timesB \to B $$
   what happens to be the type of
   $$1_A \times A \circ \lambda^{-1} f,$$ with
   $\lambda^{-1}f : A\times B \to B$
   So the diagram
   \[
   \begin{tikzcd}[column sep=normal]
     \lambda^{-1}(A \arrow{r}{1_A} & A \arrow{r}{f} & B^B) = A \times B \arrow{r}{1_A \times B} & A \times B \arrow{r}{\lambda^{-1} f} & B 
   \end{tikzcd}
   \]
   matches the second axiom.

@@

  \\
  So $\forall A,B,C \in \textbf{Cat}$, the exponential object is given as $C^B$, with the evaluation arrow
  $$\epsilon: C^B \times B \to C $$ and the transposition
  $$\lambda f: A \to C^B,\forall f:A \times B \to C$$
  The equation
  $$f = \epsilon \circ (\lambda f \times 1_B)$$ can be verified with the commuting diagram
  \[
  \begin{tikzcd}[row sep=huge]
    A \arrow{d}{\lambda f} & A \times B \arrow{d}{\lambda f \times 1_B} \arrow{dr}{f} \\
    C^B                    & C^B \times B \arrow{r}{\epsilon} & C 
  \end{tikzcd}
  \]
  \\
  For some $g:A\toC^B$, we have the inverse of the transpose
  $$\lambda^{-1}g: A\times B \to C$$ with
  $$\lambda^{-1}g:=\epsilon \circ (g\times 1_B)$$
  and need to show, that
  $$g = \lambda(\epsilon \circ (g\times 1_B).$$
  Since we already know, that $(\lambda \circ \lambda^-1) (g) = g$, there is nothing left to show and \textbf{Cat} is cartesian closed.

@@

Beyond the already seen isomorphism $0 \times A \cong 0$, many identities knwon from common arithmetics,like for instance that of the natural numbers,
hold for cartesian closed categories.$\forall A, B, C \in $ in some cartesian closed category $\mathcal{A}$ with terminal object "1" and binary products "$\times$": \\
\begin{itemize}
\item $A \times (B \times C) \cong (A \times B) \times C$
\item $A \times B \cong B \times A$
\item $ 1^A \cong 1 $
\item $ A^1 \cong A$
\item $(A \times B)^C \cong A^C \times B^C$
\item $(A^B)^C \cong A^{B \times C} $
\end{itemize}
\\
If $\mathcal{A}$ is bicartesian closed with initial object "0" and co-product "+", the enumeration of axioms above is extended by: 
\begin{itemize}
\item $A+B \cong B+A$
\item $(A+B)+C \cong A+(B+C)$
\item $(A+B)\times C \cong A \times C+B \times C$
\item $A^{B+C} \cong A^B \times A^C $
\item $0+A \cong A$
\item $A^0 \cong 1$
\end{itemize}
\\
We are going to show some of these identities by transformations exploiting right or left adjointness.
Usually it suffices to show, that both sides satisfy the same universal property.
\begin{itemize}
\item $$1^A \cong 1.$$
  holds, because there is only one arrow from every object A $\in ob(\mathcal{A})$ to the terminal object 1 which makes Hom(A,1) containing exactly one arrow. The terminal object itself is merely equipped with the identity $1_1$.
  So both sides satisfy the universal property and are therefore isomorphic.
  \\

\item $$(A^B)^C \cong A^{B \times C}.$$
  Take some X,Y $\in ob(\mathcal{A})$ and set
  $$ X:=Hom(-, (A^B)^C) \quad and \quad Y:= Hom(-,A^{B\timesC})$$ then
  $$ Hom(-, (A^B)^C) \cong Hom(-\times C, A^B) $$
  $$ \cong Hom((- \times C)\times B,A) $$
  With B, associativity and commutativity of the product
  $$ Hom(-,A^{B\times C}) \cong Hom(- \times (B\times C),A) $$
  $$ \cong Hom(-\times (C\times B),A) $$
  $$ \cong Hom((-\times C)\times B,A) $$
  So $X \cong Y$, what shows the claim.
\item $$A^0 \cong 1$$
  The argument is similar to the one above. There is only one arrow from the initial object 0 to any other object A, so Hom(0,A) is isomorphic to the terminal object 1.
\end{itemize}

@@

\\
Subcategories $\mathcal{B}$ of cartesian closed categories $\mathcal{A}$ are in general not cartesian closed.
A simple counterexample is the category of categories with finite products\textbf{Cat}, which is cartesian closed with binary products and exponentials. 
Yet $\textbf{Rel} \in ob(\textbf{Cat})$ is true, whereas \textbf{Rel} is not a CCC.

>>>>

\end{definition}
Since products and exponentials are given in the definition, the remainung task is, whether
 $$\textbf{HA}(a \land b, c) \cong \textbf{HA}(a, b \Rightarrow c)$$ holds.
\\
Heyting Algbras are partially ordered sets, so as in the example for \textbf{posets}, the monotonicity constraint must hold for exponentials $b^a= a \Rightarrow q$.
In the following, exponentation is written as $a \Rightarrow b = \neg a \lor b$, when necessary.
\\
So the monotonicity constraint imposes $$ a \Rightarrow b \leq b \Rightarrow c \ iff \ a \leq b \ and \ b \leq c.$$
Hence $$a \Rightarrow b  \leq a \Rightarrow b$$
$$ a \Rightarrow b \land a \leq b$$ by the condition $a \land b \leq c$ iff $a \leq b \Rightarrow c$, which is the evaluation arrow, and it is monotone
$$ a \Rightarrow b \land a = (\neg a \lor b) \land a$$
$$ = (\neg a \land a) \lor (b \land a)$$
$$ = 0 \lor b \land a \leq b$$ what always holds.

@@

   For the second axiom, choose $f:A\to B^B$ as and $p:=1_A$.Let $\mu := \lambda^{-1}$
   $$f \circ 1_A : A \to A \to B^B$$
   the transpose has type
   $$\lambda^{-1}: A \times B \to B $$
   what happens to be the type of
   $$1_A \times A \circ \lambda^{-1} f,$$ with
   $\lambda^{-1}f : A\times B \to B$
   So the diagram
   \[
   \begin{tikzcd}[column sep=normal]
     \lambda^{-1}(A \arrow{r}{1_A} & A \arrow{r}{f} & B^B) = A \times B \arrow{r}{1_A \times B} & A \times B \arrow{r}{\lambda^{-1} f} & B 
   \end{tikzcd}
   \]
   matches the second axiom.

@@

  So $\forall A,B,C \in \textbf{Cat}$, the exponential object is given as $C^B$, with the evaluation arrow
  $$\epsilon: C^B \times B \to C $$ and the transposition
  $$\lambda f: A \to C^B,\forall f:A \times B \to C$$
  The equation
  $$f = \epsilon \circ (\lambda f \times 1_B)$$ can be verified with the commuting diagram
  \[
  \begin{tikzcd}[row sep=huge]
    A \arrow{d}{\lambda f} & A \times B \arrow{d}{\lambda f \times 1_B} \arrow{dr}{f} \\
    C^B                    & C^B \times B \arrow{r}{\epsilon} & C 
  \end{tikzcd}
  \]
  \\
  For some $g:A \to C^B$, we have the inverse of the transpose
  $$\lambda^{-1}g: A\times B \to C$$ with
  $$\lambda^{-1}g:=\epsilon \circ (g\times 1_B)$$
  and need to show, that
  $$g = \lambda(\epsilon \circ (g\times 1_B).$$
  Since we already know, that $(\lambda \circ \lambda^-1) (g) = g$, there is nothing left to show and \textbf{Cat} is cartesian closed.

@@

Beyond the already seen isomorphism $0 \times A \cong 0$, many identities knwon from common arithmetics,like for instance that of the natural numbers,
hold for cartesian closed categories.$\forall A, B, C \in $ in some cartesian closed category $\mathcal{A}$ with terminal object "1" and binary products "$\times$": 
\begin{itemize}
\item $A \times (B \times C) \cong (A \times B) \times C$
\item $A \times B \cong B \times A$
\item $ 1^A \cong 1 $
\item $ A^1 \cong A$
\item $(A \times B)^C \cong A^C \times B^C$
\item $(A^B)^C \cong A^{B \times C} $
\end{itemize}

If $\mathcal{A}$ is bicartesian closed with initial object "0" and co-product "+", the enumeration of axioms above is extended by: 
\begin{itemize}
\item $A+B \cong B+A$
\item $(A+B)+C \cong A+(B+C)$
\item $(A+B)\times C \cong A \times C+B \times C$
\item $A^{B+C} \cong A^B \times A^C $
\item $0+A \cong A$
\item $A^0 \cong 1$
\end{itemize}

We are going to show some of these identities by transformations exploiting right or left adjointness.
Usually it suffices to show, that both sides satisfy the same universal property.
\begin{itemize}
\item $$1^A \cong 1.$$
  holds, because there is only one arrow from every object A $\in ob(\mathcal{A})$ to the terminal object 1 which makes Hom(A,1) containing exactly one arrow. The terminal object itself is merely equipped with the identity $1_1$.
  So both sides satisfy the universal property and are therefore isomorphic.

\item $$(A^B)^C \cong A^{B \times C}.$$
  Take some X,Y $\in ob(\mathcal{A})$ and set
  $$ X:=Hom(-, (A^B)^C) \ and \ Y:= Hom(-,A^{B \times C})$$
  then
  $$ Hom(-, (A^B)^C) \cong Hom(-\times C, A^B) $$
  $$ \cong Hom((- \times C)\times B,A) $$
  With B, associativity and commutativity of the product
  $$ Hom(-,A^{B\times C}) \cong Hom(- \times (B \times C),A) $$
  $$ \cong Hom(-\times (C\times B),A) $$
  $$ \cong Hom((-\times C)\times B,A) $$
  So $X \cong Y$, what shows the claim.
\item $$A^0 \cong 1$$
  The argument is similar to the one above. There is only one arrow from the initial object 0 to any other object A, so Hom(0,A) is isomorphic to the terminal object 1.
\end{itemize}

@@

Subcategories $\mathcal{B}$ of cartesian closed categories $\mathcal{A}$ are in general not cartesian closed.
A simple counterexample is the category of categories with finite products\textbf{Cat}, which is cartesian closed with binary products and exponentials. 
Yet $\textbf{Rel} \in ob(\textbf{Cat})$ is true, whereas \textbf{Rel} is not a CCC.
